Gradual Soundness: Lessons from Static Python

### 2022-01-15 pj-7.1/paper.tex
pj: sec 4 overhaul

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mid} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}

@@

\usepackage[framemethod=TikZ]{mdframed}
\newmdenv[roundcorner=2pt,linewidth=0.4pt,linecolor=Black]{spcodebox}
\usepackage{amsthm}
\usepackage{mathpartir}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

%% test suite is for counterex, not proof
%% 
%% (as opposed mech CONFUSSING let each sec explain purpose)
%% 
%% keep none None & explain why, ambiguity in stx
%% 
%% fully-sound weird, kill phrase ... say whats happening .... surface t is linter
%% ignore WF checker
%% 
%% check para may be red.
%% "The strict cast for checked dictionaries ...."
%% 
%% soundness, 1st para garbled don't need such a long reminder
%% 
%% Notable py **that we handle**
%% soundly => correctly
%% 
%% addl -> additionally (4.4)
%% 
%% "is triue" missing word
%% 
%% move 4.5 up to 4.1, start with the payoff --- found soundness bugs in a tool that's running in prod at instagram
%% 
%% rename 4.3 to "boundry type soundness" satate a lemma instead of full TS keep the focus, don't jump up
%% - make the proof case via Nom keep it short they have "detailed" proof
%% - "interested reader" can follow another direct path (edit current para2)
%% 
%% 4.0 put less emphasis on part2 of the soundness. ... it's not a threat its a boring swept-under-rug part ... maybe 4.5 is the new 4.0???
%% 
%% 4.4 redex = since no formalization of py + sp .. follow well-established path of using test suite to check conformance with **reality**
%% - cite Py \Js Jan paper, Politz getter-setter, Gardner js
%% 
%% 4.4 redex += everything in 4.1--4.3 is captured in redex, done 3 steps
%%  1 human verification, carefully spot-check python faithful, boundaries vs 4.1--4.3
%%  2 run againt SP tests .. indeed part of the reason for getting so much python is to run against an actual test suite
%%  3 b/c in redex, use pbt to verify conventional soundness
%% 
%% cut "soundness as baseline" final 2 sentences of 4.4(?) they're too weak, end with the compiler verification out-of-scope part

To validate the design of \SP{}, we have developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; in particular,
    interactions with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a small formalization
(as opposed to the Redex \emph{mechanization})
that illustrates the \SP{} gradual interaction strategy.
%% Compared to other gradual languages, these boundaries are quite restrictive---which
%% partly explains how it is able to perform well in practice~(\cref{s:eval}).
%% Indeed the boundaries are effectively a subset of those in Nom~\cite{mt-oopsla-2021}.
The second claim is justified by our Redex development and its test suite~(\cref{s:redex}).

\subsection{Surface Syntax and Types}

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptclass \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
  \\
    \spx, \spf & \defeq & \mbox{variable names}
  \end{langarray}

@@

\Cref{f:surface-types} illustrates key aspects of the \SP{} syntax and types.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (perhaps $\spobject$), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.\footnote{By contrast to
Python, in which \code{int} is a type and \code{None} is a value, our formalization uses lowercase names for
values and capitalized names for types.}
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.
Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written in a tidy normal form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
There is no function type in the model, or in \SP{} for that matter.
The only way to define a higher-order function is to use the dynamic type.

The surface types are meant to describe programs, but they are not designed
to be fully sound.
For this reason, we omit the surface typing judgment, which scans for
logical errors, and merely present the declarative syntax of surface types.
We also omit a well-formedness judgment that checks for circular definitions.

@@

\subsection{Evaluation Types, Casts, and Typing Judgment}

@@

The strict cast for checked dictionaries greatly simplifies type compatibility for \SP{}.
Values that are partly-dynamic are simply incompatible with static types.
For example, suppose that a function expects a list of numbers.
In a typical gradual language, a list of dynamic values would be a valid input to the function.
In \SP{}, a list of dynamic values is not a valid input.
Consequently, \SP{} can use an ahead-of-time strategy to enforce soundness.
Any position where the dynamic type meets a static type needs a cast.
Other meetings between types can be resolved statically.
\SP{} does not need to worry that a list of dynamic values might be masquerading
under a fully-static type.

@@

\subsection{Type Soundness}

The central question for the model is whether \SP{}'s approach to gradual typing
is sound.
No matter how an expression mixes typed and untyped code, its static
\emph{evaluation type} should be an accurate prediction about the results
of evaluation.
Using the metavariable $\spvalue$ to represent heap-allocated Python values
and $\sprred$ as the evaluation relation (which would need to maintain a heap),
an approximate soundness theorem is as follows:

\begin{theorem}[Type Soundness]
  If\ \(~\vdash \spexpr : \sptype\)
  then one of the following holds:
  \begin{itemize}
    \item
      \(\spexpr \sprred \spvalue
        \mbox{ and }
        \vdash \spvalue : \mftypeF{\sptype}
      \)
    \item
      \(\spexpr\) reduces to an allowed error
    \item
      \(\spexpr\) diverges
  \end{itemize}
\end{theorem}

As mentioned above, our argument for soundness has two parts
The question at hand is whether the boundaries to type dynamic
are adequately protected with run-time checks.
The Redex mechanization's comprehensive test suite provides the other half of the
argument; namely, that \SP{} is a sound type system for Python.

The direct path to boundary soundness involves two observations.
First, every evaluation type can be enforced with a decidable cast.
This is clearly the case for basic values such as floats.
It true for parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$
because of their strict semantics.
Second, all boundaries relate the dynamic type to an evaluation type.
There are no significant boundaries that relate a partially static type
to an evaluation type.
This is due to the strict semantics for checked types and the fact that
$\sptoptional{\sptdyn}$ normalizes to the dynamic type.

Alternatively, observe that the boundaries in \SP{} are a subset
of the possible boundaries in Nom, a gradual language that comes with a detailed
proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
Given an \SP{} program, we can derive a Nom program with
the same type boundaries by replacing every checked dict instance
with a fresh class type.
Because distinct checked dict instantiations are incompatible,
the \SP{} boundaries impose the same constraints as the translated Nom
boundaries.

\subsection{Redex Development, \SP{} Conformance}
\label{s:redex}
%% pbt for soundness?

By contrast to the small formalization presented in this section,
our Redex mechanization covers both type boundaries and a large subset of Python.
Notable Python features include module imports, loops, exception handlers, and
delete statements (for class members and attributes).
To ensure that these features are implemented correctly and soundly, the
model includes hundreds of test programs adapted from the \SP{} test suite.\footnote{
  Most of the 214 tests are direct translations of \SP{} conformance tests to
  the syntax of the model.
  A program that implements the translation is included with our Redex development.
  The other 52 tests are not directly expressible in the model; for these, we
  translated salient aspects of each test by hand.}
These tests exercise a variety of Python behaviors, thereby giving us confidence
that the model is sound for Python.
Some tests intentionally raise type errors, some raise exceptions, and the rest
compute a specific result; \SP{} agrees on these end-to-end behaviors.
We have additional applied Redex's random testing tools~\cite{kf-sfp-2009}
to validate soundness.
%% FILL how many random tests?
%% FILL what bugs did random catch?

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly.
These include primitive types~\cref{s:c-types} and strict modules~\cref{s:strict-mod}.
Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
are also omitted.
Other Python features are left out either because their \SP{} semantics is identical
to the baseline Python semantics or because they would entail a separate project~(\cref{s:impl}).
In particular, modelling type-directed optimization would require substantial
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.
Our model establishes soundness as a baseline.
Future work can use soundness as a starting point for compiler verification,
or informally leverage soundness to reason about possible optimizations.

\subsection{The Payoff: Issues Reported}

While modelling \SP{}, we found several issues with the language
and submitted them as bug reports to the \SP{} developers.
Overall, we reported 25 issues~(listed in \cref{a:github-issues}).
Four of these were critical soundness issues, one of which resulted in
a segmentation fault.
All but one of these soundness issue have been fixed.
Five issues were relatively minor; these dealt with confusing error messages
and incorrect tests.
The remaining 16 issues report bugs in language design and implementation.
The \SP{} team has acknowledged these bugs as such by applying a specific GitHub label (\code{sp-correctness}).

For example, we discovered that \SP{} handled the following function incorrectly when
we tried to model it:
%% \url{https://github.com/facebookincubator/cinder/issues/62}

\noindent{}The function expects either a string or the \code{None} value and promises to return a string.
When called with \code{None}, however, the function breaks out of the \code{while} loop
and implicitly returns \code{None}---disobeying the return type.
\SP{} had failed to account for the \code{break} and implicit return.
More concerningly, the associated test case \emph{expected} the program to type check.
Creating a formal model helped us catch this specification error.

@@

%% TODO SP side channels:
%% ?? eval,
%% the locals and globals dicts (the latter can be accessed through frame objects at the moment),
%% and mutable closure cells.

This section explains significant aspects of \SP{} that lie outside the
scope of our model.
These include sound interactions with Python code,
the bytecode language and optimizations, and dynamic Python features
that \SP{} does not type.

\subsection{Interactions with Open-World Python Code}

@@

Interactions with \emph{open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
Similarly, concrete types such as $\sptchkdict{\sptype}{\sptype}$ also
check their writes.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness

>>>>

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\bcinst}[1]{\code{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{~\mid~} %% {\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}
\newcommand{\ipara}[1]{\emph{#1}\/.}
\newcommand{\nnum}[1]{$N=#1$}
\newcommand{\totalnum}[1]{$#1$ total}

@@

\usepackage[framemethod=TikZ]{mdframed}
\newmdenv[roundcorner=2pt,linewidth=0.4pt,linecolor=Black]{spcodebox}
\usepackage{amsthm}
\usepackage{mathpartir}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
module imports, loops, exception handlers, and delete statements.
The model follows \SP{}'s approach to typing these features.

Because there is no prior formalization of Python and \SP{}, we cannot
\emph{verify} that the model matches reality.
Instead, we have applied the well-established method of testing end-to-end
conformance~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}
with a test suite.
In particular, we translated 214 tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,\footnote{For most of the 214 tests, the translation is automatic.
  A few tests required hand-pruning to remove features that the model
  does not handle (\totalnum{52}).}
which suggests that the model conforms to actual \SP{}.

Beyond the conformance suite, we have applied two further tests for correctness.
First, we used Redex's random testing tools~\cite{kf-sfp-2009} to check
type soundness on thousands of example programs (\totalnum{\mbox{FILL}}).
%% FILL how many programs exactly?
%% FILL what bugs did testing catch?
Second, we have manually inspected the model to convince ourselves
of its soundness.
Though not a proof, the inspection gives us confidence that the model,
and by extension \SP{}, is essentially sound.

\paragraph{The Payoff: Issues Reported}

The modelling process helped uncover several critical issues with the
implementation of \SP{}.
This is a very important payoff given that \SP{} is running in production at
Instagram.
Overall, we reported 25 bug reports to the \SP{} team via GitHub~(listed in
\cref{a:github-issues}).
Four of these were soundness issues, one of which could be easily exploited to raise
a segmentation fault.
All but one of these soundness bugs have been fixed.
Five issues were relatively minor; these dealt with confusing error messages
and incorrect tests.
The remaining 16 issues report bugs in language design and implementation.
The \SP{} team has acknowledged these bugs as such by applying a specific GitHub label: \lstinline{sp-correctness}.

To give one example issue, \SP{} accepted the following program whereas our model
reported a type error:
%% \url{https://github.com/facebookincubator/cinder/issues/62}

\noindent{}The function \code{f} expects either a string or the \code{None} value and promises to return a string.
When called with \code{None}, however, the function breaks out of the \code{while} loop
and implicitly returns \code{None}---disobeying the return type.
\SP{} had failed to account for the \code{break} and implicit return.
More concerningly, the associated test case \emph{expected} the program to type check.
Creating a formal model helped us catch this specification error.

\paragraph{Section Outline}

The rest of this section illustrates core aspects of our Redex model by way of
a small formalization.
\Cref{s:surface} presents a surface syntax to highlight the boundaries between
typed and untyped code.
\Cref{s:eval-types} explains how types get enforced at run-time.
\Cref{s:ts} argues that the overall approach toward boundaries is sound.

\subsection{Surface Syntax and Types}
\label{s:surface}

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
  \\ & &
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spf}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptclass \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
  \\
    \spx, \spf & \defeq & \mbox{variable names}
  \end{langarray}

@@

\Cref{f:surface-types} illustrates key aspects of the \SP{} syntax and types.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (either $\spobject$ or another class), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.

In Python, the syntax for expressions overlaps with the syntax of types.
The Python name \code{None}, for example, is both an expression and a type.
By contrast, \cref{f:surface-types} keeps the two syntaxes distinct.
Expressions use lowercase names (e.g. \code{none}) and types uses capitalized
names (\code{None}).

Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written in a tidy normal form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
There is no function type in the model, or in \SP{} for that matter.
The only way to define a higher-order function is to use the dynamic type.

\SP{} does not enforce all surface types against untyped code, only the evaluation
types presented in the next subsection~(\cref{s:eval-types}).
For this reason, we omit the surface typing judgment, which is a kind of linter
that merely scans typed code for logical errors.

@@

\subsection{Evaluation Types, Casts, and Typing Judgment}
\label{s:eval-types}

@@

%% 2022-01-15: unclear if this is helpful or redundant, hiding for now
%The strict cast for checked dictionaries greatly simplifies type compatibility for \SP{}.
%Values that are partly-dynamic are simply incompatible with static types.
%For example, suppose that a function expects a list of numbers.
%In a typical gradual language, a list of dynamic values would be a valid input to the function.
%In \SP{}, a list of dynamic values is not a valid input.
%Consequently, \SP{} can use an ahead-of-time strategy to enforce soundness.
%Any position where the dynamic type meets a static type needs a cast.
%Other meetings between types can be resolved statically.
%\SP{} does not need to worry that a list of dynamic values might be masquerading
%under a fully-static type.

@@

\subsection{Type Boundary Soundness}
\label{s:ts}

The soundness of our model is easily seen by reduction to Nom, a gradual
language that allows fine-grained interactions with untyped code and comes with
a detailed proof of soundness~\cite{mt-oopsla-2017,mt-oopsla-2021}.
Given a program written in the syntax of \cref{f:surface-types}, one can derive
a Nom program with the same type boundaries by replacing every checked
dict instance with a fresh class type.
Because distinct checked dict types are incompatible, the
original boundaries impose the same constraints as the translated Nom ones.

For interested readers, we offer the following direct argument as well.
Boundaries to less-typed code would be \emph{unsound} if an untyped
(or partially-untyped) value could enter a typed context without
a validating check.
The question is thus whether all boundaries are properly guarded.
Our answer has two parts.
First, observe that every evaluation type comes with a decidable
cast~(\cref{t:cast}).
For basic types such as floats, a tag check clearly suffices.
For parameterized types such as $\sptchkdict{\spteval_0}{\spteval_1}$,
the exact-match semantics also provides a decidable, though strict, test.
Second, observe that boundaries can only relate the dynamic type to an
evaluation type; there are no significant boundaries that relate a partially
static type to an evaluation type.
This is due to the strict semantics for checked types and the fact that
$\sptoptional{\sptdyn}$ normalizes to the dynamic type.
Therefore casts at static occurrences of the materialization rule are
enough to protect all boundaries.

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~\cref{s:c-types} strict modules~\cref{s:strict-mod},
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the type-directed optimizer~(\cref{s:optimize}).
Modelling type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.

\subsection{Interactions with Open-World Python Code}
\label{s:py-boundary}

@@

Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
Similarly, concrete types such as $\sptchkdict{\sptype}{\sptype}$ also
check their writes.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness
